The Name of the Title is Hope

### 2024-10-04 submission_JNE/source/input/2_introduction.tex
page and jne submission

Besides body and outcome augmentation, action augmentation has been defined as the case where a ``system assists the user’s action to produce the intended outcome.''~\cite{Cornelio2022-aq}. Such action augmentations can be realized purely on a software integration level, for example by an AI pair programmer. When it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of one's body. Unfortunately, such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.

\begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth]{figures/teaser.png}%{figures/bci_game.png}
    \caption{We propose an augmentation system that maintains agency. When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals. A brain-computer interface (BCI) then predicts the user to be in either an idle (red) or pre-movement (green) state. In the latter case, electrical muscle stimulation (EMS) is triggered and the user's hand is moved. Image taken with consent from participant.}
\end{figure}

Having a SoA means experiencing control of our own voluntary actions, instead of them feeling as randomly happening to us. It has been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}. Hence, a key challenge to drive the adoption of human action augmentation is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.

\subsection{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
In this paper, we present a prototype to overcome the disruption in experiencing agency during action augmentation. We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector. The augmentation system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.

@@

Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially even preceding their voluntary motor command. This augmentation moved the ring finger in accordance with the user's intention to act. We achieved the movement by leveraging electrical muscle stimulation (EMS) applied to the user's forearm flexor muscle. To get a first glimpse of how our system can maintain agency in augmented users, we conducted a user study using a mixed-methods research approach. In short, we explored whether aligning the physical impact on the user's body with their intention to move preserved their SoA.

>>>>

Besides body and outcome augmentation, action augmentation has been defined as the case where a ``system assists the user’s action to produce the intended outcome.''~\cite{Cornelio2022-aq}. Such action augmentations can be realized purely on a software integration level, for example by an AI pair programmer. When it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of one's own body. Unfortunately, such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\columnwidth]{figures/teaser.png}%{figures/bci_game.png}
    \caption{We propose an augmentation system that aligns with the user's agency. When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals. A brain-computer interface (BCI) then predicts the user to be in either an idle (red) or pre-movement (green) state. In the latter case, electrical muscle stimulation (EMS) is triggered and moves the user's hand. Image taken with consent from participant.}
    \label{fig:teaser}
\end{figure}

Having an SoA means experiencing control over our own voluntary actions, instead of them feeling as randomly happening to us, for a recent review see~\cite{Wen2022-bm}. It has been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}. Hence, a key challenge to drive the adoption of human action augmentation is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.

% \subsection*{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
With this paper, we strive towards an `agency-aligned' augmentation system. We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector, see figure~\ref{fig:teaser}. The closed-loop augmentation system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}. Ultimately, such an augmentation can cue the user's movements, increase their strength, and might also preempt their action, i.e., increase their speed.

@@

Upon predicting a \textit{pre-movement} state, the system augments the user's action, potentially even preceding their voluntary motor command. This augmentation moves the ring finger in accordance with the user's intention to act. We achieved the movement by leveraging electrical muscle stimulation (EMS) applied to the user's forearm flexor muscle. To get a first glimpse of how our system can maintain agency in augmented users, we conducted a user study using both quantitative and qualitative research methods. In short, we explored whether aligning the physical impact on the user's body with their intention to move, preserved their SoA.

### 2024-10-04 submission_JNE/source/input/3_related_work.tex
page and jne submission

\section{Related Work}
Our research draws inspiration from neuroscience, especially in assessing the SoA qualitatively and quantitatively, and from engineering work on BCIs as well as on physical user augmentation.

\subsection{Theories of Sense of Agency}
The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: When we intentionally perform an action, the brain generates sensory predictions about the action outcome. These predictions are constantly compared to the actual sensory data available through the sensory system during the execution of the action. These include continuous signals such as proprioceptive and visual monitoring of the ongoing movement as well as higher level predictions about the semantic outcome of the action~\cite{Clark2013-ah, Haggard2003-ff, Haggard2017-uv}. If no sensorimotor incongruency arises, a SoA manifests.

In the simple case of pressing a key on a piano, the finger movement is constantly compared to the predicted proprioceptive feedback. Subsequently, the tone generated by the key press is evaluated against auditory predictions. On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key~\cite{Pangratz2023-ew}. If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced.
%, see figure~\ref{fig:task_design} INTENTION).

In human-computer interaction (HCI) research, these constructs are categorized using a different terminology. Often, the term \textit{pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements. At higher levels of the cognitive hierarchy, \textit{reflective}, i.e. conscious, experience is used to refer to matching semantic predictions about action outcomes~\cite{Danry2022-xk, Cornelio2022-aq}.

As opposed to intended actions using our own body, movement augmentation hardware allows moving a user's body without their intention. Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e., exoskeletons, a user's body can be moved by applying forces to the extremities. Another possibility is to stimulate the brain directly, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS). Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle-activating nerves. Irrespective of the method applied, the user computes no predictions about the movement and its outcome in these scenarios. Thus, concerning the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}. With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- negatively impacts the experience of agency.

\subsubsection{Measuring Sense of Agency}
Both explicit and implicit methods have been developed to evaluate the sense of agency. These methods provide the basis to investigate the effects of action augmentation technology on agency experience. Explicit methods directly query participants to report their subjective experiences using questionnaires. Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}. However, in most cases such questionnaires aim at a higher-level, reflective, judgment of agency.

On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}. Seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}. Critically, this temporal compression only occurs following movements that were intended. The action outcome is mentally \textit{bound} to the intention. To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay~\cite{Barlas2018-bq}. Supplementing this behavioral phenomenon, physiological evidence can prove useful to further understand the interplay between volitional action and the SoA.

@@

Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA frequently use highly controlled `stimulus-response' paradigms. For example, scenarios where participants are instructed to tap on a touchscreen in response to a presented stimulus on the screen. Here, participants' behavior can be predicted with very high certainty to follow the presented stimulus, and estimating their reaction time is very accurate. In such controlled scenarios, the timing of an action augmentation device can be tuned to be near optimal. Hence, \textit{pre-empting} the user's motion can be designed to fall in line with their intention to move, thereby maintaining SoA. Previously, ~\citet{Kasahara2019-sk} used a reaction time task in which participants had to tap a target on screen as soon as it appeared and subsequently rate their SoA. They showed that in such a scenario, user's actions can be pre-empted and that a pre-emption of about 80 ms best preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.

This is in line with evidence from cognitive neuroscience which indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}. Hence, after this ``point of no return'' user's struggle to assign a source other than themselves to the action initiation. Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention to act. Cruciall then, a key challenge remaining is to design systems that maintain agency when user's actions are unpredictable and where the experimenter does \textit{not} have executive control over the environment. In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?

@@

Taken together, these findings demonstrated that the RP is a reliable signal preceding voluntary movement initiation and hence, is a well-suited candidate to base (neuro--)adaptive systems on.

>>>>

\section{Related Work}
Our research draws inspiration from neuroscience and from engineering work on BCIs as well as on physical user augmentation. In order to situate our findings, we briefly review the literature on SoA specifically focusing on what it means to act at one's own volition, being a passive observer, and when acting integrated with a technology.

\subsection{Theories on Sense of Agency}
The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: When we act at our own volition and intentionally perform an action, the brain generates sensory predictions about the action outcome. These predictions are constantly compared to the actual sensory data available during the execution of the action. These include continuous signals such as proprioceptive and visual monitoring of the ongoing movement as well as higher level predictions about the semantic outcome of the action~\cite{Clark2013-ah, Haggard2003-ff, Haggard2017-uv}. If no sensorimotor incongruency arises, and further, the brain attributes subjective causality over the action outcome, SoA manifests.

In the simple case of pressing a key on a piano, the finger movement is constantly compared to the predicted proprioceptive feedback. Subsequently, the tone generated by the key press is evaluated against auditory predictions. On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key, and in general aligns with the subjective goal of the keypress~\cite{Pangratz2023-ew}. If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced.
%, see figure~\ref{fig:task_design} INTENTION).

In human-computer interaction (HCI), these constructs are often categorized in slightly different terms. \textit{Pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements. At higher levels of the cognitive hierarchy, \textit{reflective}, i.e. conscious, experience refers to matching semantic predictions about action outcomes~\cite{Danry2022-xk, Cornelio2022-aq}.

In order to measure SoA, both explicit and implicit methods have been developed. Explicit methods directly query participants to report their subjective experiences using questionnaires. Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}. In most cases such questionnaires aim at a higher-level, reflective, judgment of agency.

On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}. Seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, e.g. a sound, participants mentally compress the delay~\cite{Haggard2002-sz}. In the theories original formulation, this temporal compression was assumed to only occur following movements that were intended: The action outcome is mentally \textit{bound} to the intention. To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay.

More recently it has been shown that this `temporal binding' also manifests when participants are merely a bystander witness in action-outcome scenarios~\cite{Suzuki2019-pi, Gutzeit2023-ei}. Among others, Suzuki et al.~\citep{Suzuki2019-pi} showed that temporal binding manifests as an effect of multisensory causal binding unrelated to intention or agency, e.g. binding effects were shown in scenarios where one is witnessing a replay of one's own earlier actions. Hence, it remains interesting to see how such binding manifests in \textit{integrated} user's.

% integration
As opposed to acting at one's own volition using one's own body, movement augmentation hardware allows moving a user's body even without their intention. Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e., exoskeletons, a user's body can be moved by applying forces to the extremities~\cite{Kuhn2013-ls}. Another possibility is to stimulate the brain directly~\cite{Haggard2002-sz}, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS). Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle-activating nerves. Irrespective of the method applied, these technologies allow to move a user's body without the user having generated any predictions about the movement and its outcome. However, proprioceptive, visual, and other signals indicate that one's own body is moving. Hence, re-afference signals are present without an efference command and copy. Thus, concerning the comparator model, a prediction error will arise, negatively impacting SoA~\cite{Wen2020-dk}.

% none or a decreased SoA will be observed in the case of externally controlled actions, for example by brain stimulation or when the body is moved by another person~\cite{Kuhn2013-ls}. With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- negatively impacts the experience of agency.

% \subsubsection{Measuring Sense of Agency}
% Both explicit and implicit methods have been developed to evaluate the sense of agency. These methods provide the basis to investigate the effects of action augmentation technology on agency experience.

% Explicit methods directly query participants to report their subjective experiences using questionnaires. Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}. In most cases such questionnaires aim at a higher-level, reflective, judgment of agency.

% On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}. Seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}. In the theories original formulation, this temporal compression was assumed to only occur following movements that were intended: The action outcome is mentally \textit{bound} to the intention. To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay~\cite{Barlas2018-bq}.

% % introduce temporal binding -- intentional binding as a measure frequently used in neuroscience and what is the state-of-the-art of the separation of these phenomena
% % ~\cite{Wen2020-dk} - prediction errors

% Supplementing this behavioral phenomenon, physiological evidence can prove useful to further understand the interplay between volitional action and the SoA.

@@

Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA, frequently use highly controlled `stimulus-response' paradigms. For example, scenarios where participants are instructed to tap on a touchscreen in response to a presented stimulus on the screen. Here, participants' behavior can be predicted with very high certainty to follow the presented stimulus, and estimating their reaction time is very accurate. In such controlled scenarios, the timing of an action augmentation device can be tuned to be near optimal. Hence, \textit{pre-empting} the user's motion can be designed to fall in line with their intention to move, thereby maintaining SoA. Previously, ~\citet{Kasahara2019-sk} used a reaction time task in which participants had to tap a target on screen as soon as it appeared and subsequently rate their SoA. They showed that in such a scenario, user's actions can be pre-empted and that a pre-emption of about 80 ms best preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.

This is in line with evidence from cognitive neuroscience which indicates that from around 200 ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}. After this ``point of no return'' user's struggle to assign a source other than themselves to the action initiation. Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention to act. Crucially then, a key challenge remaining is to design systems that maintain agency when user's actions are unpredictable and where the experimenter does \textit{not} have executive control over the environment. In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?

@@

Taken together, these findings demonstrated that the RP is a reliable signal preceding voluntary movement initiation and hence, is a well-suited candidate to base (neuro)--adaptive systems on.
